---
title: "FAST method"
author: "Kim Dill-McFarland"
date: "Oct 4, 2017"
output:
  html_document: 
    toc: true
    toc_float: true
  pdf_document: default
editor_options: 
  chunk_output_type: console
---
#Introduction
Convenient, reproducible, and rapid preservation of unique biological samples is pivotal to their use in microbiome analyses. As an increasing number of human longitudinal studies are looking to incorporate human microbiome data, the need for such sample collection and storage methods is high. Here, we describe the Fecal Aliquot Straw Technique (FAST) of fecal sample processing for long-term storage.

For 8 subjects (Sub.1-Sub.8), samples types are:

* Straw 1 (S1): straw 1 sample resulting from FAST method
* Straw 2 (S1): gavage of straw 2 sample used in humanization of mice
* Mouse fecals (F1-3): mouse fecal samples from mice humanized with different donor Subjects (N = 3 per Subject)
    + Collectively referenced as mice (M)

###

#Data
##Load packages
```{r Packages, warning=FALSE, message=FALSE}
#Manipulation of tables
library(tidyr)
library(reshape2)
#Community ecology package (Bray-Curtis, Jaccard)
library(vegan)
#Community ecology package (UniFrac)
library(phyloseq)
#Linear and nonlinear mixed effects models
library(nlme)
#Pairwise comparisons within nlme models
library(multcomp)
#Plots and visualization
library(ggplot2)
#Plotting multiple ggplot2 plots in one frame
library(cowplot)
#Venn diagrams
library(venneuler)
```

##Load functions
```{r}
#Pairwise PERMANOVAs
source("Functions/Pairwise_adonis_all.r")

#Function to identify taxa shared by S1 and S1 or S2 and M for each subject. Taxon must be at least 0.1% relaive abundance in at least one sample in the comparison for each subject.
#Creates input for mean.counts
source("Functions/venn.abund.all.r")
#Function to calculate percentages of taxa recovered across FAST (counts from venn.abund.all) and means and standard errors of these percentages
#Fig. 2B & 3B
source("Functions/mean.counts.r")

#Function to identify taxa shared by S1 and S1 or S2 and M for each subject. No abundance cutoff
#Creates input for venn.perc
source("Functions/venn.all.r")
#Function to pull shared OTUs (determined by venn.all function) out of OTU table and calculate percent relative abundance of those shared OTUs
#Fig. 2C & 3D
source("Functions/venn.perc.r")

#To obtain lists of beta-diversity values of mouse samples compared to samples from their matched subject donor (DONOR) or compared to any other subject in the dataset (OTHER).
source("Functions/donor.other.r")
```

##Load data
```{r Load.data}
#Un-rarified OTU table from mothur
OTU = read.table("Data/WLS.mouse.final.opti_mcc.unique_list.shared", sep="\t", header=TRUE)
  row.names(OTU) = OTU$Group
  OTU = OTU[,!names(OTU) %in% c("label","Group","numOtus")]

#Rarified OTU table from mothur
OTU.norm = read.table("Data/WLS.mouse.final.opti_mcc.unique_list.0.03.norm.shared", sep="\t", header=TRUE)
  row.names(OTU.norm) = OTU.norm$Group
  OTU.norm = OTU.norm[,!names(OTU.norm) %in% c("label","Group","numOtus")]

#Metadata and alpha-diversity measures  
meta = read.table("Data/Metadata.txt", sep="\t", header=TRUE, row.names=1)

#Taxonomy of OTUs from mothur
tax = read.table("Data/WLS.mouse.final.opti_mcc.unique_list.0.03.cons.taxonomy", sep="\t", header=TRUE)
  row.names(tax) = tax$OTU
  tax = separate(tax, Taxonomy, into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Strain"), sep=";")
  tax = tax[,!(names(tax) %in% c("Size", "Strain", "OTU"))]
  
#Taxa tables: OTUs summed at difference taxonomic levels
phyla = as.data.frame(t(read.table("Data/phyla_unrare.txt", sep="\t", header=TRUE, row.names=1)))
order = as.data.frame(t(read.table("Data/order_unrare.txt", sep="\t", header=TRUE, row.names=1)))
family = as.data.frame(t(read.table("Data/family_unrare.txt", sep="\t", header=TRUE, row.names=1)))
genus = as.data.frame(t(read.table("Data/genus_unrare.txt", sep="\t", header=TRUE, row.names=1)))
```

Order the data so that all tables have rows in the same order
```{r}
#Order all by samples name
OTU = OTU[order(row.names(OTU)),]
OTU.norm = OTU.norm[order(row.names(OTU.norm)),]
meta = meta[order(row.names(meta)),]
phyla = phyla[order(row.names(phyla)),]
order = order[order(row.names(order)),]
family = family[order(row.names(family)),]
genus = genus[order(row.names(genus)),]
```

Calculate OTU precent relative abundance
```{r}
abund=OTU/rowSums(OTU)*100
```

###

#Data manipulation
##Phyloseq object
All samples
```{r}
#Load rarified data into phyloseq objects
OTU.UF = otu_table(as.matrix(OTU.norm), taxa_are_rows=FALSE)
Tax.UF = tax_table(as.matrix(tax))
meta.UF = sample_data(meta)

#Load pre-calculated neighbor-joining tree
load("Data/NJ.tree.Rdata")

#Merge into phyloseq object
physeq.tree = merge_phyloseq(OTU.UF, Tax.UF, meta.UF, NJ.tree)
physeq.tree
```

Mouse-derived samples only
```{r}
OTU.norm.M = OTU.norm[meta$Type=="F",]
OTU.UF.M = otu_table(as.matrix(OTU.norm.M), taxa_are_rows=FALSE)
meta.UF.M = sample_data(meta[meta$Type=="F",])

physeq.tree.M = merge_phyloseq(OTU.UF.M, Tax.UF, meta.UF.M, NJ.tree)
physeq.tree.M
```

Human-derived samples only
```{r}
OTU.norm.H = OTU.norm[meta$Type !="F",]
OTU.UF.H = otu_table(as.matrix(OTU.norm.H), taxa_are_rows=FALSE)
meta.UF.H = sample_data(meta[meta$Type!="F",])

physeq.tree.H = merge_phyloseq(OTU.UF.H, Tax.UF, meta.UF.H, NJ.tree)
physeq.tree.H
```

##Calculate beta-diversity
All samples
```{r}
set.seed(35426)
#Bray-Curtis
BC.dist=as.matrix(vegdist(OTU.norm, distance="bray"))
#Jaccard
J.dist=(BC.dist*2)/(BC.dist+1)
#Weighted UniFrac
wUF.dist = as.matrix(UniFrac(physeq.tree, weighted=TRUE, normalized=TRUE))
#Unweighted UniFrac
uwUF.dist = as.matrix(UniFrac(physeq.tree, weighted=FALSE, normalized=TRUE))
```

##Calculate nMDS
For use in Fig. 2A, 3A, & S3
```{r}
set.seed(35426)
#All samples
uwUF.nmds = ordinate(physeq.tree, method="NMDS", distance="unifrac", weighted=FALSE)
#Human-derived samples
uwUF.nmds.H = ordinate(physeq.tree.H, method="NMDS", distance="unifrac", weighted=FALSE)
#Mouse-derived samples
uwUF.nmds.M = ordinate(physeq.tree.M, method="NMDS", distance="unifrac", weighted=FALSE)
```

##Count taxa recover
For use in Fig. 2B & 3B

Taxa shared by S1 and S1 *or* S2 and M for each subject. Taxon must be at least 0.1% relaive abundance in at least one sample in the comparison for each subject.

Count and taxa lists at each taxonomic level
```{r}
venn.abund.all(OTU, "OTU")
venn.abund.all(genus, "genus")
venn.abund.all(family, "family")
venn.abund.all(order, "order")
venn.abund.all(phyla, "phyla")
```

Read in results tables
```{r}
taxa.OTU = read.table("output_venn.abund.all/OTU.venn.abund.all.count.csv", sep=",", header=FALSE, row.names=1)
  colnames(taxa.OTU) = c("count","comp","subject")
taxa.genus = read.table("output_venn.abund.all/genus.venn.abund.all.count.csv", sep=",", header=FALSE, row.names=1)
  colnames(taxa.genus) = c("count","comp","subject")
taxa.family = read.table("output_venn.abund.all/family.venn.abund.all.count.csv", sep=",", header=FALSE, row.names=1)
  colnames(taxa.family) = c("count","comp","subject")
taxa.order = read.table("output_venn.abund.all/order.venn.abund.all.count.csv", sep=",", header=FALSE, row.names=1)
  colnames(taxa.order) = c("count","comp","subject")
taxa.phyla = read.table("output_venn.abund.all/phyla.venn.abund.all.count.csv", sep=",", header=FALSE, row.names=1)
  colnames(taxa.phyla) = c("count","comp","subject")
```

Convert counts to percentages of taxa recovered across FAST and means and standard errors of these percentages.

Run across taxonomic levels
```{r}
mean.counts(c("taxa.OTU", "taxa.genus", "taxa.family", "taxa.order", "taxa.phyla"))
```

Read in results
```{r}
taxa.recover = read.table("output_mean.counts/mean.counts.csv", sep=",", header=FALSE, row.names=1) 
  colnames(taxa.recover) = c("mean","se","comp","taxa")
#Add variable of taxa in order  
  taxa.recover$taxa.ord = ifelse(taxa.recover$taxa == "taxa.OTU", "5.OTU",
                          ifelse(taxa.recover$taxa == "taxa.genus", "4.genus",
                          ifelse(taxa.recover$taxa == "taxa.family", "3.family",
                          ifelse(taxa.recover$taxa == "taxa.order", "2.order",
                          ifelse(taxa.recover$taxa == "taxa.phyla", "1.phyla", NA)))))
```

##Relative abundance taxa recover
For Fig. 2C & 3D

How much of straw1 relative abundance occurs in straw2?  
How much of straw2 relative abundance occurs in mice?

Identify taxa shared by S1 and S1 or S2 and M for each subject. No abundance cutoff.
```{r}
venn.all(OTU, "OTU")
```

Read in results
```{r}
OTU.list = read.table("output_venn.all/OTU.venn.all.list.csv", sep=",", row.names=1)
  colnames(OTU.list) = c("OTU","comparison","Subject")
```

Pull shared OTUs (determined by venn.all) out of OTU table and calculate percent relative abundance of those shared OTUs
```{r}
venn.perc(OTU)
```

Read in results
```{r}
venn.perc = read.csv("output_venn.perc/venn.perc.csv", header=FALSE, row.names=1)
  colnames(venn.perc) = c("Percent","Comparison","Group","Subject")
```

##Bloom/lost in mice
Identify taxa that are more or less abundant in mouse samples compared to respective donor samples.
```{r}
#Calculate relative abundance of genera
genus.abund = genus/rowSums(genus)*100
#Remove S1 samples as comparison is S2 to M
genus.abund = genus.abund[!meta$Type == "S1",]
#Remove genera at <1% relative abundance across all samples
genus.abund.cutoff = genus.abund[, apply(genus.abund, MARGIN=2, function(x) any(x > 1))]

#Melt data from short/wide to long/skinny for all genera (columns)
genus.meta = merge(meta, genus.abund.cutoff, by="row.names")
    genus.meta = genus.meta[,!colnames(genus.meta) %in% c("Row.names", "chao", "shannon")]

#Create variable for samples in desired order    
genus.meta$Type.ord = ifelse(genus.meta$Type == "S2", "1_S2",
                  ifelse(genus.meta$Type == "F", "2_F", genus.meta$Type))
genus.meta$Type.ord = as.factor(genus.meta$Type.ord)

#Melt metadata from short/wide to long/skinny
genus.melt = melt(genus.meta, id.vars=c("Type","HM","Subject","Sample", "Type.ord"))
```

##Donor vs. other
For Fig. 3E

To obtain lists of beta-diversity values of mouse samples compared to samples from their matched subject donor (DONOR) or compared to any other subject in the dataset (OTHER).

Rename subjects to numbers so that distance may be calculated
```{r}
meta$Subject.num = ifelse(meta$Subject == "Sub.1",1,
                    ifelse(meta$Subject == "Sub.2",2,
                     ifelse(meta$Subject == "Sub.3",3,
                      ifelse(meta$Subject == "Sub.4",4,
                       ifelse(meta$Subject == "Sub.5",5,
                         ifelse(meta$Subject == "Sub.6",6,                                               ifelse(meta$Subject == "Sub.7",7,
                             ifelse(meta$Subject == "Sub.8",8,NA))))))))
```                                             

Pull out DONOR and OTHER beta-values
```{r}
donor.other(meta$Subject.num, meta$Type, c("BC","J","uwUF","wUF"))
```

Read in results
```{r}
donor.other = read.csv("output_donor.other/donor.other.csv", header=FALSE, row.names=1)
  colnames(donor.other) = c("distance","comparison","metric")
```

###

#Statistics
##Beta-diversity
###Mouse fecals
Do the mouse fecal samples differ within each individual?
```{r}
adonis(BC.dist[row.names(meta[meta$Type == "F",]),row.names(meta[meta$Type == "F",])] ~ Sample, data=meta[meta$Type == "F",], permutations=1000, strata=meta[meta$Type == "F",]$Subject)

adonis(J.dist[row.names(meta[meta$Type == "F",]),row.names(meta[meta$Type == "F",])] ~ Sample, data=meta[meta$Type == "F",], permutations=1000, strata=meta[meta$Type == "F",]$Subject)

adonis(wUF.dist[row.names(meta[meta$Type == "F",]),row.names(meta[meta$Type == "F",])] ~ Sample, data=meta[meta$Type == "F",], permutations=1000, strata=meta[meta$Type == "F",]$Subject)

adonis(uwUF.dist[row.names(meta[meta$Type == "F",]),row.names(meta[meta$Type == "F",])] ~ Sample, data=meta[meta$Type == "F",], permutations=1000, strata=meta[meta$Type == "F",]$Subject)
```

Do mice innoculated from different donors differ from one another?
```{r}
adonis(BC.dist[row.names(meta[meta$Type == "F",]),row.names(meta[meta$Type == "F",])] ~ Subject, data=meta[meta$Type == "F",], permutations=1000)

adonis(J.dist[row.names(meta[meta$Type == "F",]),row.names(meta[meta$Type == "F",])] ~ Subject, data=meta[meta$Type == "F",], permutations=1000)

adonis(wUF.dist[row.names(meta[meta$Type == "F",]),row.names(meta[meta$Type == "F",])] ~ Subject, data=meta[meta$Type == "F",], permutations=1000)

adonis(uwUF.dist[row.names(meta[meta$Type == "F",]),row.names(meta[meta$Type == "F",])] ~ Subject, data=meta[meta$Type == "F",], permutations=1000)
```

###Mouse vs human
Do human- (S1, S2) and mouse-derived (F) samples differ overall, regardless of subject?
```{r}
adonis(BC.dist ~ HM, data=meta, permutations=1000)
adonis(J.dist ~ HM, data=meta, permutations=1000)
adonis(wUF.dist ~ HM, data=meta, permutations=1000)
adonis(uwUF.dist ~ HM, data=meta, permutations=1000)
```

###Full model
Do samples (S1, S2, F) differ within subjects?
```{r}
adonis(BC.dist ~ Type, data=meta, permutations=1000, strata=meta$Subject)
adonis(J.dist ~ Type, data=meta, permutations=1000, strata=meta$Subject)
adonis(wUF.dist ~ Type, data=meta, permutations=1000, strata=meta$Subject)
adonis(wUF.dist ~ Type, data=meta, permutations=1000, strata=meta$Subject)
```

Pairwise between all sample types
```{r}
pairwise.adonis.all(OTU.norm, meta, meta$Type, "bray", NJ.tree, stratify=TRUE, meta$Subject, "bonferroni")

pairwise.adonis.all(OTU.norm, meta, meta$Type, "jaccard", NJ.tree, stratify=TRUE, meta$Subject, "bonferroni")

pairwise.adonis.all(OTU.norm, meta, meta$Type, "wunifrac", NJ.tree, stratify=TRUE, meta$Subject, "bonferroni")

pairwise.adonis.all(OTU.norm, meta, meta$Type, "uwunifrac", NJ.tree, stratify=TRUE, meta$Subject, "bonferroni")
```

###Donor vs. other
Is a mouse more like it's human donor than other humans in the dataset?
```{r}
kruskal.test(donor.other[donor.other$metric == "BC",]$distance ~ donor.other[donor.other$metric == "BC",]$comparison)

kruskal.test(donor.other[donor.other$metric == "J",]$distance ~ donor.other[donor.other$metric == "J",]$comparison)

kruskal.test(donor.other[donor.other$metric == "wUF",]$distance ~ donor.other[donor.other$metric == "wUF",]$comparison)

kruskal.test(donor.other[donor.other$metric == "uwUF",]$distance ~ donor.other[donor.other$metric == "uwUF",]$comparison)
```

##Alpha-diversity
###Mouse fecals
Do the mouse fecal samples differ within each individual?
```{r}
summary(aov(meta[meta$Type == "F",]$shannon ~ Sample + Error(Subject/Sample), data=meta[meta$Type == "F",]))
summary(aov(meta[meta$Type == "F",]$chao ~ Sample + Error(Subject/Sample), data=meta[meta$Type == "F",]))
```

Do mice innoculated from different donors differ?
```{r}
summary(aov(meta[meta$Type == "F",]$shannon ~ Subject, data=meta[meta$Type == "F",]))
summary(aov(meta[meta$Type == "F",]$chao ~ Subject, data=meta[meta$Type == "F",]))
```

###Mouse vs human
Do human- (S1, S2) and mouse-derived (F) samples differ overall, regardless of subject?
```{r}
summary(aov(shannon ~ HM, data=meta))
summary(aov(chao ~ HM, data=meta))
```

###Full model
Do samples (S1, S2, F) differ within subjects?
```{r}
shan.mod = lme(shannon ~ Type, random = ~1|Subject, data=meta)
anova(shan.mod)
#Pairwise
summary(glht(shan.mod,linfct=mcp(Type="Tukey")))

chao.mod = lme(chao ~ Type, random = ~1|Subject, data=meta)
anova(chao.mod)
#Pairwise
summary(glht(chao.mod,linfct=mcp(Type="Tukey")))
```

#Figures
##Figure 2A
Non-metric multidimensional scaling plots of the unweighted UniFrac metric between human-derived samples, colored by subject. Colored ellipses are smallest area for human samples from each subject drawn in Illustrator. 

```{r}
nMDS_uwUF_H=plot_ordination(physeq.tree.H, uwUF.nmds.H, type="sites", color="Subject", shape="Type") + 
  theme_bw() + 
  scale_shape_manual(values=c(17,15,3)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_rect(color="black", fill=NA)) +
  ggtitle("Unweighted UniFrac")

#View
nMDS_uwUF_H

#Save
save_plot("Figures/Fig2A.eps", nMDS_uwUF_H)
```

##Figure 2B
Taxa recovery in S2 compared to S1. Calculated with non-rarified data. Identified taxa were at least 0.1% relative abundance in at least one of the two samples in each comparison. 

```{r}
S12.recovery=ggplot(taxa.recover[taxa.recover$comp == "S12",], aes(taxa.ord, mean)) + 
  geom_bar(fill = "darkorange", position = "dodge", stat="identity") +
  geom_errorbar(aes(ymax=taxa.recover[taxa.recover$comp == "S12",]$mean +
  taxa.recover[taxa.recover$comp == "S12",]$se, ymin=taxa.recover[taxa.recover$comp == "S12",]$mean - taxa.recover[taxa.recover$comp == "S12",]$se)) +
  theme(legend.position="none")

#View
S12.recovery
#Save
save_plot("Figures/Fig2B.eps", S12.recovery, base_height=3, base_width=4)
```

##Figure 2C
Percent relative abundance of S1 community captured in oral gavage preparation (S2). Where shown, bars represent mean ± standard error.

```{r}
perc.retain.S12 = ggplot(venn.perc[venn.perc$Group %in% c("S1S2"),], aes(x=Group, y=Percent)) + 
  geom_boxplot(aes(group=Group)) +
  geom_jitter(width=0.15, height=0, aes(color=Subject)) +
  ylim(98,100)

#View
perc.retain.S12
#Save
postscript("Figures/Fig2C.ps", width = 3, height = 3, horizontal = FALSE, colormodel = "rgb", fonts=c("serif"))
perc.retain.S12
dev.off()
```

##Figure 3A
Non-metric multidimensional scaling plots of the unweighted UniFrac metric between mouse fecal samples, colored by donor. Colored ellipses are smallest area for mouse-derived samples from each subject drawn in Illustrator. 

```{r}
nMDS_uwUF_M=plot_ordination(physeq.tree.M, uwUF.nmds.M, type="sites", color="Subject", shape="Type") + 
  theme_bw() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_rect(color="black", fill=NA)) +
  ggtitle("Unweighted UniFrac") 

#View
nMDS_uwUF_M

#Save
save_plot("Figures/Fig3A.eps", nMDS_uwUF_M)
```

##Figure 3B
Taxa recovery in mouse fecal samples compared to oral inoculum (S2). Calculated with non-rarified data. Identified taxa were at least 0.1% relative abundance in at least one of the four samples in each comparison (S2 and 3 mouse fecal samples). 

```{r}
S2M.recovery=ggplot(taxa.recover[taxa.recover$comp == "S2M",], aes(taxa.ord, mean)) +   
  geom_bar(fill = "darkgreen", position = "dodge", stat="identity")+
  geom_errorbar(aes(ymax=taxa.recover[taxa.recover$comp == "S2M",]$mean +
  taxa.recover[taxa.recover$comp == "S2M",]$se, ymin=taxa.recover[taxa.recover$comp == "S2M",]$mean - taxa.recover[taxa.recover$comp == "S2M",]$se)) +
    theme(legend.position="none")

#View
S2M.recovery
#Save
save_plot("Figures/Fig3B.eps", S2M.recovery, base_height=3, base_width=4)
```

##Figure 3C
OTU Venn diagrams. Yellow circles represent OTUs in the oral inoculum (A=S2) and blue circles represent OTUs recovered in the mouse (B=M).

Create venns scaled to total OTU counts from `venn.abund.all` function. Cutoff of 0.1% in at least one sample in a given venn.
```{r}
Sub.1_S2M = venneuler(c('A'=13,'A&B'=61,'B'=3))
Sub.2_S2M = venneuler(c('A'=23,'A&B'=90,'B'=0))
Sub.3_S2M = venneuler(c('A'=27,'A&B'=73,'B'=8))
Sub.4_S2M = venneuler(c('A'=22,'A&B'=83,'B'=5))
Sub.5_S2M = venneuler(c('A'=23,'A&B'=79,'B'=14))
Sub.6_S2M = venneuler(c('A'=20,'A&B'=74,'B'=1))
Sub.7_S2M = venneuler(c('A'=17,'A&B'=74,'B'=1))
Sub.8_S2M = venneuler(c('A'=44,'A&B'=89,'B'=12))
```

Save Venns in 1 figure
```{r}
pdf("Figures/Fig.3C.pdf", width = 20, height = 8, colormodel = "rgb", fonts=c("serif"))
 
layout(matrix(c(1,2,3,4,5,6,7,8), 2, 4))
plot(Sub.1_S2M, col=c("yellow","blue"))
plot(Sub.2_S2M, col=c("yellow","blue"))
plot(Sub.3_S2M, col=c("yellow","blue"))
plot(Sub.4_S2M, col=c("yellow","blue"))
plot(Sub.5_S2M, col=c("yellow","blue"))
plot(Sub.6_S2M, col=c("yellow","blue"))
plot(Sub.7_S2M, col=c("yellow","blue"))
plot(Sub.8_S2M, col=c("yellow","blue"))

dev.off()
```

##Figure 3D
Percent relative abundance of the oral gavage community (S2) captured in mouse fecal samples.

```{r}
perc.retain.M = ggplot(venn.perc[venn.perc$Group %in% c("S2M"),], aes(x=Group, y=Percent)) + 
  geom_boxplot(aes(group=Group)) +
  geom_jitter(width=0.15, height=0, aes(color=Subject)) +
  ylim(55,100)

#View
perc.retain.M
#Save
postscript("Figures/Fig3D.ps", width = 3, height = 3, horizontal = FALSE, colormodel = "rgb", fonts=c("serif"))
perc.retain.M
dev.off()
```

##Figure 3E
Bray-Curtis (BC), Jaccard (J), weighted UniFrac (wUF), and unweighted UniFrac (uwUF) beta-diversity measures of mouse samples compared to samples from their matched subject donor (DONOR) or compared to any other subject in the dataset (OTHER).

Pull out distances from donor.other for each comparison and metric
```{r}
match.BC = as.vector(donor.other[donor.other$comparison == "match" & donor.other$metric == "BC",]$distance)
unmatch.BC = as.vector(donor.other[donor.other$comparison == "unmatch" & donor.other$metric == "BC",]$distance)

match.J = as.vector(donor.other[donor.other$comparison == "match" & donor.other$metric == "J",]$distance)
unmatch.J = as.vector(donor.other[donor.other$comparison == "unmatch" & donor.other$metric == "J",]$distance)

match.wUF = as.vector(donor.other[donor.other$comparison == "match" & donor.other$metric == "wUF",]$distance)
unmatch.wUF = as.vector(donor.other[donor.other$comparison == "unmatch" & donor.other$metric == "wUF",]$distance)

match.uwUF = as.vector(donor.other[donor.other$comparison == "match" & donor.other$metric == "uwUF",]$distance)
unmatch.uwUF = as.vector(donor.other[donor.other$comparison == "unmatch" & donor.other$metric == "uwUF",]$distance)
```

Calculate mean and standard errors
```{r}
dist.summary = data.frame(
  mean = c(mean(match.BC), mean(unmatch.BC),
           mean(match.J), mean(unmatch.J),
           mean(match.wUF), mean(unmatch.wUF),
           mean(match.uwUF), mean(unmatch.uwUF)),
  se = c(sd(match.BC)/sqrt(length(match.BC)),
         sd(unmatch.BC)/sqrt(length(unmatch.BC)),
         sd(match.J)/sqrt(length(match.J)),
         sd(unmatch.J)/sqrt(length(unmatch.J)),
         sd(match.wUF)/sqrt(length(match.wUF)),
         sd(unmatch.wUF)/sqrt(length(unmatch.wUF)),
         sd(match.uwUF)/sqrt(length(match.uwUF)),
         sd(unmatch.uwUF)/sqrt(length(unmatch.uwUF))),
  comp = rep(c("match", "unmatch"), times=4),
  dist = c("BC","BC","J","J","wUF","wUF","uwUF","uwUF"),
  group = c("1mBC","2uBC","3mJ","4uJ","5mwUF","6uwUF","7muwUF","8uuwUF"))
```

Plot means with se error bars
```{r}
#Set limits of error bars to be +/- standard error
limits = aes(ymax = dist.summary$mean + dist.summary$se, ymin = dist.summary$mean - dist.summary$se)

#Create base plot
p1 = ggplot(data = dist.summary, aes(x = group, y = mean, color=comp))

#Add data bars
bar_donor_other=
  p1 + geom_bar(fill='white', stat = "identity", position = position_dodge(0.9)) +
  geom_errorbar(limits, position = position_dodge(0.9),  width = 0.25) +
  labs(x = "", y = "Distance") + 
  theme_bw() + 
  theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), axis.text = element_text(colour="black"), axis.ticks = element_line(colour="black")) +
  scale_color_manual(values=c('black','black'))  +
    theme(legend.position="none")
  
#View
bar_donor_other

#Save
save_plot("Figures/Fig3E.eps", bar_donor_other, base_height=2.5, base_width=3.5)
```

##Figure S1
Shannon’s diversity of microbial communities for the eight sequenced human samples. S1 (red) and S2 (yellow) represent randomly selected straws for each subject. 

```{r}
#Create variable of subject wih sample type
meta$Subject.type = paste(meta$Subject, meta$Type)

shannon.plot = ggplot(meta[meta$Type %in% c("S1","S2"),], aes(Subject.type, shannon)) +   
  geom_bar(aes(fill = Type), position = "dodge", stat="identity") +
  ggtitle("Shannon") +
  scale_fill_manual(values=c("red","yellow"))

#View
shannon.plot
#Save
save_plot("Figures/FigS1.eps", shannon.plot, base_height=3, base_width=4)
```

##Figure S2
Genera that tended to be lost or bloomed in humanized mice. Relative abundances of genera at least 1% relative abundance in at least one straw2 (S2) or mouse (M) sample. Colored by subject and split into (A) more highly abundant and (B) lower abundance genera. Bars for mice are averages across the three mouse fecal samples for each subject.

First, explore all taxa
```{r}
bloom.taxa.all=ggplot(genus.melt, aes(variable:Type.ord, value)) +   
  geom_bar(aes(fill = Subject), position = "dodge", stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

bloom.taxa.all
```

Select those with visual pattern of higher or lower in mice than respective donor.
```{r}
#Fig S2A - Higher abundance taxa
bloom.taxa.high=ggplot(genus.melt[genus.melt$variable %in% c("g__Akkermansia","g__Bacteroides","g__Ruminococcus"),], aes(variable:Type.ord, value)) +   
  geom_bar(aes(fill = Subject), position = "dodge", stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#Fig S2B - Lower abundance taxa
bloom.taxa.low=ggplot(genus.melt[genus.melt$variable %in% c("g__Bifidobacterium","g__Collinsella","g__Dialister","g__Lachnospira","g__Streptococcus","g__Sutterella"),], aes(variable:Type.ord, value)) +   
  geom_bar(aes(fill = Subject), position = "dodge", stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

bloom.taxa.high
bloom.taxa.low

postscript("Figures/FigS2A.ps", width = 3, height = 4, horizontal = FALSE, colormodel = "rgb", fonts=c("serif"))
bloom.taxa.high
dev.off()

postscript("Figures/FigS2B.ps", width = 6, height = 4, horizontal = FALSE, colormodel = "rgb", fonts=c("serif"))
bloom.taxa.low
dev.off()
```

##Figure S4
Unweighted UniFrac beta-diversity of FAST and mouse fecal microbiota. Non-metric multidimensional scaling plots of the unweighted UniFrac metric between all samples. Samples are colored by subject. Shapes indicate sample type. Black ellipses are standard error for human and mouse groups.

```{r}
nMDS_uwUF_all=plot_ordination(physeq.tree, uwUF.nmds, type="sites", color="Subject", shape="Type") + 
  theme_bw() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_rect(color="black", fill=NA)) +
  ggtitle("Unweighted UniFrac") +
  stat_ellipse(aes(group=HM), level=0.95)

#View
nMDS_uwUF_all

#Save
save_plot("Figures/FigS4.eps", nMDS_uwUF_all)
```

FIN